---
output:
  html_document: default
  pdf_document: default
---
# Microbiome-Environment Spatial Analysis Pipeline

**A comprehensive workflow for modeling microbial diversity using real environmental data and spatial cross-validation**

---

## Overview

This analysis demonstrates a complete pipeline for modeling microbial diversity patterns across landscapes using real environmental data layers. The workflow incorporates:

### 🌍 **Real Environmental Data Integration**
- WorldClim climate layers (temperature, precipitation)
- SoilGrids soil properties (pH, organic carbon)
- Proper spatial resolution matching and resampling

### 🔬 **Realistic Microbiome Simulation**
- Environment-driven community composition
- Multinomial count data with realistic sequencing depths
- Phyloseq integration for standard microbiome workflows

### 📊 **Spatial Statistics**
- Moran's I test for spatial autocorrelation in diversity
- Mantel test for distance-decay relationships
- Conditional spatial cross-validation implementation

### 🤖 **Machine Learning Best Practices**
- XGBoost with hyperparameter tuning
- Spatially-aware cross-validation when needed
- Overfitting diagnostics and early stopping
- Variable importance analysis

### 🗺️ **Prediction and Visualization**
- Large-scale spatial prediction mapping
- Interactive leaflet visualizations
- Model validation against environmental drivers

---

## Technical Skills Demonstrated

- **R Programming**: Advanced data manipulation, statistical analysis, and visualization
- **Spatial Analysis**: Raster processing, coordinate systems, spatial statistics
- **Machine Learning**: XGBoost, cross-validation, model diagnostics
- **Microbiome Analytics**: Phyloseq, diversity metrics, community ecology
- **Data Visualization**: Static plots, interactive maps, diagnostic plots
- **Reproducible Research**: Structured workflow, documentation, version control ready

---

The analysis focuses on Spain, simulating microbiome data on Catalonia that responds realistically to environmental gradients to predict alpha diversity of the country.

---

## Setup and Dependencies

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = FALSE
)
```

```{r libraries, message=FALSE}
# ===========================
# MICROBIOME + ENVIRONMENT PIPELINE (REAL ENV LAYERS + SOILGRIDS + SPATIAL CV)
# ===========================
library(phyloseq)
library(vegan)
library(tidyverse)
library(sf)
library(terra)
library(geodata)
library(xgboost)
library(caret)
library(leaflet)
library(leafem)
library(geodist)
library(rnaturalearth)
library(rnaturalearthhires)
library(spdep)
library(blockCV)
library(geosphere)
library(patchwork) 

set.seed(42069)
```

---

## 1. Study Area and Sample Design

We generate random sampling locations across Catalonia, Spain to simulate a realistic field sampling campaign.

```{r sample-design}
# ===========================
# 1) Simulate Metadata (Cataluña)
# ===========================
n_samples <- 100
spain_states <- ne_states(country = "Spain", returnclass = "sf")
catalonia    <- spain_states[spain_states$woe_name == "Cataluña", ]

random_points <- st_sample(catalonia, size = n_samples)
metadata <- st_coordinates(random_points) %>% as.data.frame()
metadata$SampleID <- paste0("S", 1:n_samples)
colnames(metadata) <- c("Longitude", "Latitude", "SampleID")
rownames(metadata) <- metadata$SampleID

cat("Generated", n_samples, "sampling locations across Catalonia\n")
```

---

## 2. Environmental Data Acquisition

We fetch real environmental data from two major global databases:

### 2.1 WorldClim Climate Data
- **BIO1**: Annual Mean Temperature (°C)
- **BIO12**: Annual Precipitation (mm)

### 2.2 SoilGrids Soil Properties
- **pH**: Soil pH (H2O) at 5-15cm depth
- **OC**: Organic Carbon Density at 5-15cm depth

```{r environmental-data}
# ===========================
# 2) Fetch Real Env Data
# ===========================
# 2a) WorldClim (Temp & Precip)
worldclim <- geodata::worldclim_global(var = "bio", res = 5, path = "worldclim_data")
wc_layers <- worldclim[[c(1, 12)]] # BIO1 (== Temp) & BIO12 (== Precip)
names(wc_layers) <- c("Temp", "Precip")

# 2b) SoilGrids (pH & Organic Carbon)
sg_layers <- geodata::soil_world(var = c("phh2o","ocd"), path="soilgrids_data", depth = 15, res = 5) ### THIS PULLS THE SOILGRIDS RASTERS (organic carbon and pH)
sg_layers <- sg_layers[[c("phh2o_5-15cm", "ocd_5-15cm")]]
names(sg_layers) <- c("pH","OC")

# NOTE: for whatever reason, soil_world ignores res = 5, so i need to resample to match worldclim's and be able to merge. fine.
res(wc_layers)
res(sg_layers)
sg_resampled <- terra::resample(sg_layers, wc_layers, method = "bilinear")

# Combine all layers
env_layers <- c(wc_layers, sg_resampled)

# Extract for sample points
pts_vect  <- vect(st_as_sf(metadata, coords=c("Longitude","Latitude"), crs=4326))
env_data  <- terra::extract(env_layers, pts_vect)[,-1]
metadata  <- bind_cols(metadata, env_data)

# Remove samples with NA (probaly landed on marine areas where there's no soil ans such)
valid_rows <- complete.cases(metadata)
metadata   <- metadata[valid_rows, ]
pts_vect   <- pts_vect[valid_rows, ]
env_data   <- env_data[valid_rows, ]

# update n_samples
n_samples <- nrow(metadata)

cat("Successfully extracted environmental data for", n_samples, "samples\n")
cat("Environmental layers:", names(env_layers), "\n")
```

---

## 3. Microbiome Data Simulation

We simulate realistic microbiome count data where community composition is driven by environmental gradients, specifically pH in this case.

```{r microbiome-simulation}
# ===========================
# 3) Simulate Microbiome Counts (driven by real env)
# ===========================
n_otus <- 1000

# base abundances
base_ab  <- rlnorm(n_otus, meanlog=4, sdlog=1)

# pick variable!
colnames(metadata)
top_Var <- c("pH")
var_distro <- 0.4

# OTU responses
top_Var_effect  <- rnorm(n_otus,0,var_distro) # strong effect. to soften, 0.1 or less

# expected abundances
expected_abund <- t(sapply(1:n_samples, function(j){
  ba <- base_ab * exp(top_Var_effect     * metadata[[top_Var]][j])
  ba / sum(ba)
}))

colnames(expected_abund) <- paste0("OTU",1:n_otus)
rownames(expected_abund) <- metadata$SampleID

# simulate sequencing depths and counts
seq_depth <- rpois(n_samples, lambda=20000)
otu_mat   <- sapply(1:n_samples, function(i){
  rmultinom(1, size=seq_depth[i], prob=expected_abund[i,])
})
otu_mat   <- matrix(otu_mat, nrow=n_otus, dimnames=list(paste0("OTU",1:n_otus), metadata$SampleID))

head(otu_mat)

# MERGE METADATA AND MICROBIOME DATA TO CREATE THE PHYLOSEQ OBJECT
ps <- phyloseq(
  otu_table(otu_mat, taxa_are_rows = TRUE),
  sample_data(metadata)
)

cat("Created phyloseq object with", n_otus, "OTUs and", n_samples, "samples\n")
cat("Environmental driver selected as primary:", top_Var,"\n")
```

---

## 4. Alpha Diversity Calculation

We calculate Shannon and Simpson diversity indices for each sample.

```{r diversity-calculation}
# ===========================
# 4) EStimate richness
# ===========================
div <- estimate_richness(ps, measures=c("Shannon","Simpson")) %>%
  rownames_to_column("SampleID")
div_data <- left_join(metadata, div, by="SampleID")

head(div_data)

cat("Shannon diversity range:", round(range(div_data$Shannon, na.rm = TRUE), 2), "\n")
cat("Simpson diversity range:", round(range(div_data$Simpson, na.rm = TRUE), 3), "\n")
```

---

## 5. Spatial Autocorrelation Analysis

Before modeling, we test for spatial autocorrelation in our diversity data using two complementary approaches:

### 5.1 Moran's I Test
Tests for spatial autocorrelation in Shannon diversity values based on geographic proximity.

### 5.2 Mantel Test
Tests for correlation between community dissimilarity and geographic distance.

```{r spatial-autocorrelation}
# ===========================
# 5) Spatial Autocorrelation Tests
# ===========================
# MORAN (on Shannon's div)
coords     <- cbind(div_data$Longitude, div_data$Latitude)
nb         <- spdep::knn2nb(spdep::knearneigh(coords, k=5))
lw         <- spdep::nb2listw(nb, style="W")
mor_test   <- spdep::moran.test(div_data$Shannon, lw)
print(mor_test)

# MANTEL 
comm_dist <- phyloseq::distance(ps, method = "bray")
coords <- cbind(div_data$Longitude, div_data$Latitude)
geo_dist <- geosphere::distm(coords, fun = distHaversine) #unit is in meters
geo_dist <- as.dist(geo_dist)
mantel_test <- vegan::mantel(comm_dist, geo_dist, method = "pearson", permutations = 9999)
print(mantel_test)

# If significant autocorrelation by either moran or mantel, set up 5 spatial CV blocks
if (mor_test$p.value < 0.05 | mantel_test$signif < 0.05){
  
  AUTOCORR = TRUE # for later ;)
  
  sf_data <- st_as_sf(div_data, coords = c("Longitude", "Latitude"), crs = 4326)
  
png("spatial_blocks.png", width = 800, height = 800, res = 150)

blocks <- cv_spatial(
  x = sf_data,
  column = NULL,
  k = 5,
  size = 50000,   # block size in meters (50 km)
  selection = "random",
  seed = 420,
  progress = FALSE
)

dev.off()

plot(blocks)
  
  cat("Spatial autocorrelation detected: using spatial cross-validation\n")
  
} else{
  
  AUTOCORR = FALSE
  
  print("No spatial autocorrelation found.")
  
}
```


---

## 6. Machine Learning Model Development

We use XGBoost to predict Shannon diversity from environmental variables, employing spatial cross-validation if autocorrelation is detected.

### 6.1 Model Training and Cross-Validation

```{r model-training}
# ===========================
# 4) XGBoost Model (Spatially Aware if Needed)
# ===========================
model_data <- div_data %>% drop_na()

# Check if we have enough data
if (nrow(model_data) < 10) {
  stop("Not enough valid data points for modeling. Check environmental data extraction.")
}

# Features and response
X <- as.matrix(model_data[, c("Temp", "Precip", "pH", "OC")])
y <- model_data$Shannon

dtrain <- xgb.DMatrix(data = X, label = y)

# define folds
if (AUTOCORR) {
  print("Created spatial folds.")
  folds <- split(seq_along(y), blocks$folds_ids)
} else {
  print("Created regular folds.")
  folds <- caret::createFolds(y, k = 5)
}

# XGBoost Cross-Validation
params <- list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,
  max_depth = 3,
  subsample = 0.5,
  colsample_bytree = 0.5
)

set.seed(420)
cv <- xgb.cv(
  params = params,
  data = dtrain,
  folds = folds,  # spatial or random folds
  nrounds = 1000,
  early_stopping_rounds = 20,
  verbose = 0
)

cat("Cross-validation completed with", cv$best_iteration, "optimal rounds\n")
```

### 6.2 Model Diagnostics and Overfitting Assessment

```{r model-diagnostics, fig.width=12, fig.height=4}
# Overfitting Diagnostics
best_iter <- cv$best_iteration
eval_df <- cv$evaluation_log

# Plot 1: Train vs Test RMSE Curves
p1 <- ggplot(eval_df, aes(x = iter)) +
  geom_line(aes(y = test_rmse_mean, color = "Test"), size = 1) +
  geom_line(aes(y = train_rmse_mean, color = "Train"), size = 1) +
  geom_vline(xintercept = best_iter, color = "darkgreen", linetype = "dashed", size = 1) +
  annotate("text", x = best_iter, 
           y = eval_df$test_rmse_mean[best_iter] + 0.02,
           label = paste0("Gap: ", 
                          round(eval_df$train_rmse_mean[best_iter] -
                                eval_df$test_rmse_mean[best_iter], 4)),
           color = "black", vjust = -0.5, size = 3.5) +
  scale_color_manual(values = c("Test" = "red", "Train" = "blue")) +
  labs(y = "RMSE", x = "Iterations", color = "Curve") +
  theme_minimal() +
  ggtitle("Train vs Test RMSE")

# Plot 2: RMSE Gap
eval_df$rmse_gap <- eval_df$train_rmse_mean - eval_df$test_rmse_mean

p2 <- ggplot(eval_df, aes(x = iter, y = rmse_gap)) +
  geom_line(color = "purple", size = 1) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  labs(y = "Train - Test RMSE", x = "Iterations") +
  theme_minimal() +
  ggtitle("RMSE Gap Across Iterations")

combined_rmse <- p1 + p2 + plot_layout(ncol = 2)

print(combined_rmse)

ggsave("xgboost_rmse_plots.png", combined_rmse, width = 12, height = 6, dpi = 300)

cat("Overfitting gap at best iteration:", round(eval_df$rmse_gap[best_iter], 4), "\n")
```

### 6.3 Final Model Training and Evaluation

```{r final-model}
# ============================
# Train Final Model (using best nrounds)
# ============================
best_nrounds <- cv$best_iteration

set.seed(42069)
xgb_model <- xgboost(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,
  verbose = 0
)

# Predictions & Evaluation (Train-Test Split) /random split only for evaluation metrics (not used for CV)
set.seed(420)
train_idx <- caret::createDataPartition(y, p = 0.7, list = FALSE)
X_test  <- X[-train_idx, ]
y_test  <- y[-train_idx]

pred <- predict(xgb_model, newdata = X_test)
correlation <- cor(y_test, pred)
cat("Model performance (Shannon ~ predicted):", round(correlation, 6), "\n")

# Variable Importance
importance <- xgb.importance(model = xgb_model)
par(mfrow = c(1, 1))
print(importance)
```

```{r importance-plot, fig.width=8, fig.height=5}
xgb.plot.importance(importance)
```

---

## 7. Spatial Prediction and Mapping

We apply our trained model to predict Shannon diversity across the entire Spanish landscape.

```{r spatial-prediction}
# ===========================
# 6) Predict Over Spain Raster
# ===========================
# prepare raster, aggregate for speed
spain_sf   <- ne_countries(country="Spain", returnclass="sf")
env_crop   <- crop(env_layers, ext(spain_sf)) %>% mask(vect(spain_sf))
env_agg    <- aggregate(env_crop, fact=2, fun=mean, na.rm=TRUE)

# predict function
pred_fun <- function(model, newdata){
  mat <- as.matrix(data.frame(Temp=newdata[,1],
                              Precip=newdata[,2],
                              pH=newdata[,3],
                              OC=newdata[,4]))
  predict(model, mat)
}

# Pedict over raster using the trained XGBoost model
predicted_map <- terra::predict(
  env_agg, xgb_model,
  fun = pred_fun,
  na.rm = TRUE
)

# Set layer name
names(predicted_map) <- "Shannon_Predicted"

cat("Successfully generated predictions across Spain\n")
```

### 7.1 Model Validation: Prediction vs. Environmental Driver

```{r validation-plot, fig.width=12, fig.height=5}
## compare to variable used to create microbiome data. should be highly correlated!
spain_map <- crop(env_layers[[top_Var]], predicted_map)
spain_map <- resample(spain_map, predicted_map, method = "bilinear")
spain_map <- mask(spain_map, predicted_map)

# Convert rasters to data frames for ggplot
pred_df <- as.data.frame(predicted_map, xy = TRUE)
colnames(pred_df) <- c("LON", "LAT", "Shannon")

spain_df <- as.data.frame(spain_map, xy = TRUE)
colnames(spain_df) <- c("LON", "LAT", "Var")

# Predicted Shannon Diversity
p1 <- ggplot() +
  geom_raster(data = pred_df, aes(x = LON, y = LAT, fill = Shannon)) +
  scale_fill_viridis_c(option = "viridis", name = "Shannon") +
  geom_point(data = div_data, aes(x = Longitude, y = Latitude),
             color = "red", size = 1.5) +
  theme_minimal() +
  ggtitle("Predicted Shannon Diversity")

# Environmental Driver (Top Variable)
p2 <- ggplot() +
  geom_raster(data = spain_df, aes(x = LON, y = LAT, fill = Var)) +
  scale_fill_viridis_c(option = "viridis", name = top_Var) +
  theme_minimal() +
  ggtitle(paste("Environmental Driver:", top_Var))


combined_plot <- p1 + p2 + plot_layout(ncol = 2)

ggsave("diversity_environmental_driver.png", combined_plot,
       width = 14, height = 6, dpi = 300)

print(combined_plot)

cat("Validation: Comparing predictions to", top_Var, "(the primary environmental driver). Should be highly correlated! \n")
```

---

## 8. Interactive Visualization

```{r interactive-map, fig.width=10, fig.height=8}
# ===========================
# 7) Interactive Map & Summary
# ===========================
# leaflet
pal <- colorNumeric(terrain.colors(50), values(predicted_map), na.color="transparent")
leaflet() %>%
  addTiles() %>%
  addRasterImage(predicted_map, colors=pal, opacity=0.7, group="Predicted Shannon") %>%
  addCircleMarkers(data=st_as_sf(div_data, coords=c("Longitude","Latitude"),crs=4326),
                   color="red", radius=5,
                   popup=~paste0("Samp:", SampleID,
                                 "<br>Shannon:",round(Shannon,2),
                                 "<br>Temp:",round(Temp,1),
                                 "<br>Precip:",round(Precip,0),
                                 "<br>pH:",round(pH,2),
                                 "<br>OC:",round(OC,1)),
                   group="Samples") %>%
  addLegend(pal=pal, values=values(predicted_map), title="Predicted Shannon", position="bottomright")
```